Redundant calibration as a complex optimization problem

### 2017-11-12 paperG2.tex
added the almost final version of the paper

The second advantage of redundant array configurations is that they, by definition, they measure the same spatial scales {\bf (GB: perhaps worth having the Van cittert-Zernike theorem here?)} regardless of what the sky brightness distribution looks like. This means that the corruptions that are introduces by the instrument can be calibrated - up to some extent - without knowing anything about the sky brightness distribution \citep[][]{Noordam1982,Wieringa1992}. This property is particularly desirable for EoR observations as it mitigates errors introduced by an incomplete knowledge of the sky brightness distribution.

In this paper we extend the stefcal calibration algorithm to the redundant use case using the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
We also perform a proper computational analysis of the recently proposed (preconditioned conjugate gradient) PCG method when it is applied to the redundant calibration problem.
We also compare these two approaches.

present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. With respect to current algorithms, it is more robust to initial conditions, it is comparatively fast and {\bf ?????...}. We also present an analysis of its computational complexity showing that the LM algorithm shows better convergence properties than the GN currently used \cite{Zheng14}.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015,Kurien2016} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we test the algorithm performances in Section~\ref{sec:results} and we conclude in Section\ref{sec:conclusions}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is given therefore by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

@@

We can now formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T, &  &\label{eq:parm_definitions}
 \end{align}
where the number of model parameters to be solved of is now $P = 2(N+L)$, since redundant calibration is a complex problem.
%The number of model parameters that we need to solve for in equation~\ref{eq:least_squares_red} is $P = 2(N+L)$, since redundant calibration is a complex problem.

% \subsection{Limitations}
% \label{sec:scope}
% We only deal with per-timeslot and frequency channel calibration. No solution smoothing  
% is applied between the different timeslots and channels. The reader interested in this is referred to \citep{Zheng2014}. 
% Moreover, we do not explicitly deal with any of the degeneracies normally associated with redundant calibration \citep{Zheng2014,Kurien2016}. Similar to the approach taken in \citep{Marthi2014}, we assume that the degeneracies present in the redundant calibration solutions are removed with a final round of skymodel-based calibration.

As mentioned in Section~\ref{sec:sky_wirtinger}, equation~\ref{eq:least_squares_red} can be solved by splitting the problem into its real and
imaginary parts and then solve for the real and imaginary parts of the unknown parameters separately \citep{Wieringa1992,Liu2010,Zheng2014}. In Appendix~\ref{sec:lincal} we review this approach.
%As already alluded to in Section~\ref{sec:sky_wirtinger}, we can solve equation~\ref{eq:least_squares_red} in one of two ways. We can split the problem into its real and imaginary parts and then solve for the real and imaginary parts of the unknown parameters separately or we can employ the notion of Wirtinger derivatives. The former approach is discussed in Section~\ref{sec:ri}, while the latter approach is described in Section~\ref{sec:w}. 
%
%\subsection{Real and Imaginary Approach}
%\label{sec:ri}
%Using least-squares to solve complex problems can be problematic as $\bv$ needs to be analytic in its argument. In the case of equation~\ref{eq:least_squares_red}, \citet{Liu2010} circumvents this difficulty by splitting the problem into its phase-and-amplitude components.
%
%The aforementioned is achieved by first rewriting equation~\ref{eq:least_squares_red} as
%\begin{equation}
%\label{eq:least_squares_lincal}
%\min_{\bmath{\varrho}} \|\breve{\br}\| = \min_{\bmath{\varrho}} \|\breve{\bd} - \breve{\bv}(\bmath{\varrho})\|, 
%\end{equation}
%where $\breve{\bd} = [\bd^T,\conj{\bd}^T]^T$, $\breve{\bv} = [\bv^T,\conj{\bv}^T]^T$ and $\bmath{\varrho}$ is the phase and amplitude parameter vector which we define in Appendix~\ref{sec:lincal}. 
%With the aid of equation~\ref{eq:least_squares_lincal} we can derive the algorithm implemented in \textsc{lincal} (see Appendix~\ref{sec:lincal}). The Jacobian associated with equation~\ref{eq:least_squares_lincal} is equal to equation~\ref{eq:Jac_lin}. Appendix~\ref{sec:lincal} makes it evident that \textsc{lincal} is equivalent to a basic GN update step \citep{Kurien2016}. 
%Therefore, a straightforward way of improving the convergence properties of stock standard \textsc{lincal} would be to simply add the LM damping factor to the basic \textsc{lincal} update step.
%In doing so the need to use $\textsc{logcal}$ before $\textsc{lincal}$ becomes superfluous. The use of $\textsc{logcal}$ may, however, still offer a computational advantage, as it is a very fast method to obtain a good starting solution, and would reduce the number of iterations needed by the LM algorithm. Proper experiments would need to be conducted to determine whether it is worth keeping $\textsc{logcal}$ in the algorithmic hierarchy. 
%
%\subsection{Wirtinger Approach}
%\label{sec:w}
%
%\citet{Smirnov2015} recently proposed to use the Wirtinger derivative instead of the classic real and imaginary approach to perform normal skymodel-based calibration. We will now show
%that the Wirtinger derivative can also be applied to redundant calibration.
%We can perform redundant calibration by solving the following optimization problem: 
%\begin{equation}
%\label{eq:least_squares_red}
%\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
%\end{equation}
%where $\bz = [\bg,\by]^T$. The other vectors in equation~\ref{eq:least_squares_red} were defined in Section~\ref{} \footnote{In this paper we use the same symbol to denote corrupted predicted visibilities and true observed corrupted visibilities as it
%is trivial to distinguish which one is implied from the context. Using the same symbol for these two quantities improves the readability of the paper.}. Note the similarity between equation~\ref{eq:least_squares} and equation~\ref{eq:least_squares_red},
%i.e. redundant calibration can be cast as a non-linear least squares problem.

@@

%{\bf (GB: derived in which way? It would e useful to say in order to differentiate from us)}.

@@

>>>>

The second advantage of redundant array configurations is that they, by definition, they measure the same spatial scales

{\bf (GB: perhaps worth having the Van cittert-Zernike theorem here?)}

{\bf (TLG: Leave the decision to Gianni.)}

regardless of what the sky brightness distribution looks like. This means that the corruptions that are introduces by the instrument can be calibrated - up to some extent - without knowing anything about the sky brightness distribution \citep[][]{Noordam1982,Wieringa1992}. This property is particularly desirable for EoR observations as it mitigates errors introduced by an incomplete knowledge of the sky brightness distribution.

%In this paper we extend the stefcal calibration algorithm to the redundant use case using the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
%We also perform a proper computational analysis of the recently proposed (preconditioned conjugate gradient) PCG method when it is applied to the redundant calibration problem.
%We also compare these two approaches.

{\bf (TLG: I have only modified the paragraph below in the introduction to try and point out the point of the paper. I hope this makes 
more sense now given the the restructuring GB suggested.)}

In this paper we present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
With respect to current algorithms, it is more robust to initial conditions, while remaining comparatively fast. We also show 
that given certain approximations this new algorithm reduces to the redundant calibration equivalent of the StEfCal algorithm \citep{Salvini2014}.
In this paper we will refer to this simplified version of the new algorithm as redundant StEfCal. We also 
investigate the speed-up that the preconditioned conjugate gradient (PCG) method provides if it is employed by the new algorithm \citep{Liu2010}.
A comparison between the computational complexity of the optimized new algorithm and redundant StEfCal is also performed.

{\bf (TLG: The following papers on redundant calibration has to be cited in the paper \citep{Noorishad2012} and Sievers2017. It was previously given in the 
intro and was subsequently removed in the previous iteration made by GB. Unsure were they now belong.)}

%Redundant StEfCal is much faster 
%than WRC. 
%We also investigate the speed-up that is achievable by employing
%Moreover, a proper computational complexity comparison is made between redundant StEfCal and the recently proposed  when applied to the redundant calibration problem is conducted.
%With respect to current algorithms, it is more robust to initial conditions, it is comparatively fast and {\bf ?????...}. 
%We also compare these two approaches.
%We also present an analysis of its computational complexity showing that the LM algorithm shows better convergence properties than the GN currently used \cite{Zheng14}.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we test the algorithm performances in Section~\ref{sec:results} and we conclude in Section\ref{sec:conclusions}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

{\bf (GB: Trienko, I am not sure about where to place the paragraph below till the end of the subsection... probably because I'm not sure what ``even" and ``odd" \textsc{StEfCal} steps are... can we add the GN step after equation~\ref{eq:stef_lambda}???)}

{\bf (TLG: I have rewritten the above to make the odd and even stefcal steps more clear. I have also move the GN simplified equation before the 
stefcal explanation. GB to rewrite.)}

@@

We can now formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T, &  &\label{eq:parm_definitions}
 \end{align}
where the number of model parameters to be solved of is now $P = 2(N+L)$, since redundant calibration is a complex problem.
{\bf (GB: is the following important? I thought it could be removed...) We plot $L$ and $P$ for two different geometric layouts in Fig~\ref{fig:pl}.
The analytic expressions for the red and the blue curve in Fig.~\ref{fig:ln} are \citep{Camps2003}
\begin{equation}
L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2} 
\end{equation}
and
\begin{equation}
L = N - 1 
\end{equation}
respectively.}

{\bf (TLG: Done.)}

%The number of model parameters that we need to solve for in equation~\ref{eq:least_squares_red} is $P = 2(N+L)$, since redundant calibration is a complex problem.

{\bf (TLG: This section got removed during a previous edit. It needs to be put back somewhere????)}

\subsection{Limitations}
 \label{sec:scope}
 We only deal with per-timeslot and frequency channel calibration. No solution smoothing  
 is applied between the different timeslots and channels. The reader interested in this is referred to \citep{Zheng2014}. 
 Moreover, we do not explicitly deal with any of the degeneracies normally associated with redundant calibration \citep{Zheng2014,Kurien2016}. Similar to the approach taken in \citep{Marthi2014}, we assume that the degeneracies present in the redundant calibration solutions are removed with a final round of skymodel-based calibration.

As mentioned in Section~\ref{sec:sky_wirtinger}, equation~\ref{eq:least_squares_red} can be solved by splitting the problem into its real and
imaginary parts and then solve for the real and imaginary parts of the unknown parameters separately \citep{Wieringa1992,Liu2010,Zheng2014}. In Appendix~\ref{sec:logcal} and~\ref{sec:lincal} we review this approach.
%As already alluded to in Section~\ref{sec:sky_wirtinger}, we can solve equation~\ref{eq:least_squares_red} in one of two ways. We can split the problem into its real and imaginary parts and then solve for the real and imaginary parts of the unknown parameters separately or we can employ the notion of Wirtinger derivatives. The former approach is discussed in Section~\ref{sec:ri}, while the latter approach is described in Section~\ref{sec:w}. 
%
%\subsection{Real and Imaginary Approach}
%\label{sec:ri}
%Using least-squares to solve complex problems can be problematic as $\bv$ needs to be analytic in its argument. In the case of equation~\ref{eq:least_squares_red}, \citet{Liu2010} circumvents this difficulty by splitting the problem into its phase-and-amplitude components.
%
%The aforementioned is achieved by first rewriting equation~\ref{eq:least_squares_red} as
%\begin{equation}
%\label{eq:least_squares_lincal}
%\min_{\bmath{\varrho}} \|\breve{\br}\| = \min_{\bmath{\varrho}} \|\breve{\bd} - \breve{\bv}(\bmath{\varrho})\|, 
%\end{equation}
%where $\breve{\bd} = [\bd^T,\conj{\bd}^T]^T$, $\breve{\bv} = [\bv^T,\conj{\bv}^T]^T$ and $\bmath{\varrho}$ is the phase and amplitude parameter vector which we define in Appendix~\ref{sec:lincal}. 
%With the aid of equation~\ref{eq:least_squares_lincal} we can derive the algorithm implemented in \textsc{lincal} (see Appendix~\ref{sec:lincal}). The Jacobian associated with equation~\ref{eq:least_squares_lincal} is equal to equation~\ref{eq:Jac_lin}. Appendix~\ref{sec:lincal} makes it evident that \textsc{lincal} is equivalent to a basic GN update step \citep{Kurien2016}. 
%Therefore, a straightforward way of improving the convergence properties of stock standard \textsc{lincal} would be to simply add the LM damping factor to the basic \textsc{lincal} update step.
%In doing so the need to use $\textsc{logcal}$ before $\textsc{lincal}$ becomes superfluous. The use of $\textsc{logcal}$ may, however, still offer a computational advantage, as it is a very fast method to obtain a good starting solution, and would reduce the number of iterations needed by the LM algorithm. Proper experiments would need to be conducted to determine whether it is worth keeping $\textsc{logcal}$ in the algorithmic hierarchy. 
%
%\subsection{Wirtinger Approach}
%\label{sec:w}
%
%\citet{Smirnov2015} recently proposed to use the Wirtinger derivative instead of the classic real and imaginary approach to perform normal skymodel-based calibration. We will now show
%that the Wirtinger derivative can also be applied to redundant calibration.
%We can perform redundant calibration by solving the following optimization problem: 
%\begin{equation}
%\label{eq:least_squares_red}
%\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
%\end{equation}
%where $\bz = [\bg,\by]^T$. The other vectors in equation~\ref{eq:least_squares_red} were defined in Section~\ref{} \footnote{In this paper we use the same symbol to denote corrupted predicted visibilities and true observed corrupted visibilities as it
%is trivial to distinguish which one is implied from the context. Using the same symbol for these two quantities improves the readability of the paper.}. Note the similarity between equation~\ref{eq:least_squares} and equation~\ref{eq:least_squares_red},
%i.e. redundant calibration can be cast as a non-linear least squares problem.

@@

{\bf (GB: derived in which way? It would be useful to say in order to differentiate from us)}.
{\bf (TLG: Added)}.

@@

{\bf (GB: is this true? TRIENKO, I MANAGED TO REACH ONLY TILL THIS POINT. I THINK IT IS TIME I HAND THIS BACK TO YOU AND YOU MAKE AN EFFORT TO TIE IT UP. i THINK WE CAN GO FROM HERE TO A MINOR REVISION NEXT TIME)}
{\bf (TLG: tried to rephrase is it clear now?}
